---
title: "Differential expression using DESeq2"
output: html_document
---


Update: 14th February 2020


# Differential gene expression using DESeq2
Differential gene expression using DESeq2 and visualisation with Venn diagrams. Using Adam's code, some from https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

If need to use Biomart to map gene symbols from Ensembl IDs, then use functions_biomart.R instead of functions.R

```{r, include=FALSE}
source('functions.R')
library(org.Hs.eg.db)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(DT)
library(limma)
library(pheatmap)
library(dplyr)
```


```{r read count and metadata}

df_mRNA <- read.delim("GSE80774_MV411.RNAseq_pool.htseq.txt.gz", sep = "\t", header = TRUE, row.names = 1)

meta_data <- read.table("meta_data.csv", sep=",", header = TRUE)

rownames(meta_data) <- meta_data$Sample

df_mRNA = df_mRNA[,rownames(meta_data)]

all(rownames(meta_data) %in% colnames(df_mRNA))


```


```{r metadata}
# set up meta_data for all experiments (only have 1 comparison/metadata sheet)

meta_data_1 <- meta_data %>% 
  filter(Cell.line == "MV411") %>% 
  column_to_rownames("Sample")

#meta_data_2 <- meta_data %>%
  #filter (Cell_line =="Other") %>%
  #column_to_rownames("Sample")



```




```{r DESeq}
## Set up the DESeq2 object

run_deseq2 <- function(df_mRNA, meta_data, control="shNeg", test="shENL", value){

  df_mRNA = df_mRNA[,rownames(meta_data)]
  
  
  dds<- DESeqDataSetFromMatrix(countData=df_mRNA,
                               colData=meta_data,
                               design= ~Condition)
  
  keep <- rowSums(counts(dds)) >= 10
  dds <- dds[keep,]
  
  dds <- DESeq(dds)
  
  res <- results(dds, contrast = c(value, test,control))
  
  return(res)
}

de_res <- run_deseq2(as.data.frame(df_mRNA), meta_data_1, control = "shNeg", test="shENL", value="Condition")

#Keep code below if running multiple comparisons

#Too many objects called res later, so this "master" results is de_res (differential expression results. Obviously)

# stim_res <- run_deseq2(as.data.frame(gene_counts), meta_data_1, control = "soluble", test="stim", value="Condition")



```

The following describes the analysis of the data using DEseq2. I have separated the analyses into the four different contrasts and have plotted summary statistics and MA plots for each.  


## Summary of the data {.tabset .tabset-fade}

### shNeg vs shMLLT1

```{r summary}
summary(de_res, alpha=0.05)
```


## MA (mean average) plots {.tabset .tabset-fade}

In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points will be colored red if the adjusted p value is less than 0.01. Points which fall out of the window are plotted as open triangles pointing either up or down.

### shNeg vs shMLLT1

```{r MA plot}
DESeq2::plotMA(de_res)
```


# Volcano plots {.tabset .tabset-fade}

### shNeg vs shMLLT1

```{r volcano}

library("ggplot2") #Best plots
library("ggrepel") #Avoid overlapping labels


test <- as.data.frame(de_res)
  
data <- as.vector(rownames(test))
annots <-  AnnotationDbi::select(org.Hs.eg.db, keys=data,
                                   columns="SYMBOL", keytype = "ENSEMBL")
  
result <- merge(test, annots, by.x="row.names", by.y="ENSEMBL")
res <- result %>% 
    dplyr::select(log2FoldChange, SYMBOL, baseMean, padj, Row.names) %>% 
    na.omit()
  

mutateddf <- mutate(res, sig=ifelse(res$padj<0.05, "padj<0.05", "Not Sig")) #Will have different colors depending on significance
input <- cbind(gene=rownames(mutateddf), mutateddf )
input <- input %>% 
  arrange(input$padj)

#symbol_data <- head(input, 100)
#symbol_data <- input %>% 
  #filter(SYMBOL == "CD69" | SYMBOL == "IL2RA" | SYMBOL == "IFNG" | SYMBOL == "IL2")

#convert the rownames to a column
volc = ggplot(input, aes(log2FoldChange, -log10(padj))) + #volcanoplot with log2Foldchange versus pvalue
    geom_point(aes(col=sig)) + #add points colored by significance
#geom_point(data=symbol_data, aes(log2FoldChange, -log10(padj)), colour="red") +
      ggtitle("shNeg vs shMLLT1") #e.g. 'Volcanoplot DESeq2'
#volc+geom_text_repel(data=symbol_data, aes(label=`SYMBOL`)) + scale_colour_Publication() + theme_bw()#adding text for the genes

volc


#Save Volcano Plot
ggsave("volcano/shNeg_vs_shMLLT1.jpeg", device="jpeg") #In case you want to easily save to disk


```



## Results tables

The folowing results tables show the significant genes. Filtering has been performed with a log2 fold change +/- 2.

### shNeg vs shMLLT1

```{r significant genes}


dt <- filter_genes(as.data.frame(de_res), name="shNeg_vs_shMLLT1")




datatable(dt$sig)


```



