---
title: "Enrichment"
output: html_document
---

# Gene ontology and gene set enrichment analysis
Pathway analysis using code from Rbio Session 4. Gene ontology, network analysis and gene set enrichment analysis using ClusterProfiler.

Update: 14th February 2020

Note: used simplify() for every analysis - if editing and want to Knit to HTML, will take a long time!


```{r, include=FALSE}

library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(pathview)
library(fgsea)
library(readr)

#Go to Rbio.Rmd file to load full list, including installing packages etc. If having trouble, install individual packages using CRAN within R Studio. If still having trouble, then have to search Bioconductor for it.

```
Read in significant genes only (p<0.05)
```{r}

sig_genes <- read_csv("results/shNeg_vs_shMLLT1_sig.csv")


```

# Gene Ontology


Filter data
```{r}

#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
sig_genes$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(sig_genes$Row.names), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
sig_genes <- sig_genes[is.na(sig_genes$Entrez.Gene)==FALSE,]

#Remove duplicated entries
sig_genes <- sig_genes[!duplicated(sig_genes$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test - Felce Upregulated genes, Biological Process

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(sig_genes$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

#dimensions - number of GO terms
dim(ego)


#Export list of biological processes and genes involved
sig_genes_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(sig_genes_BP, file="GO/shMLLT1_BP.csv")

simp_ego <- simplify(ego)

dim(simp_ego)

#Export list of biological processes and genes involved
simp_BP <- data.frame(simp_ego$ID, simp_ego$Description, simp_ego$p.adjust, simp_ego$geneID)

write.csv(simp_BP, file="GO/shMLLT1_simp_BP.csv")

```
##All GO terms

```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```

```{r, fig.height=6, fig.width=8}
emapplot(ego, showCategory=10)

```

```{r, fig.height=14, fig.width=16}
#logFC is optional, but if available will colour-code the nodes
logFC <- sig_genes$log2FoldChange
#set name of object
names(logFC) <- sig_genes$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 10, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```

##Simplified GO terms
```{r,fig.height=6, fig.width=8}
barplot (simp_ego, showCategory = 10)

dotplot (simp_ego, showCategory=10)

```

```{r, fig.height=6, fig.width=8}
emapplot(simp_ego, showCategory=20)

```

```{r, fig.height=14, fig.width=16}
#logFC is optional, but if available will colour-code the nodes
logFC <- sig_genes$log2FoldChange
#set name of object
names(logFC) <- sig_genes$Entrez.Gene
head(logFC)

cnetplot(simp_ego, foldChange = logFC, showCategory = 6, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```
