---
title: "mcf_go"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

# Gene ontology and gene set enrichment analysis for shared MCF-7 genes between CRISPR and SGC-iMLLT RNA-Seq
Pathway analysis using code from Rbio Session 4. Gene ontology using cluster Profiler https://yulab-smu.github.io/clusterProfiler-book/


```{r, include=FALSE}
library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
#Go to Rbio.Rmd file to load full list, including installing packages etc. If having trouble, install individual packages using CRAN within R Studio. If still having trouble, then have to search Bioconductor for it.
```

## 24 hours downregulated

```{r}
library(readr)
genes_24h <- read_csv("mcf_shared_genes_24h_down.csv")
genes_24h[1] <- NULL

# Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
genes_24h$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(genes_24h$SYMBOL), keytype="SYMBOL", column="ENTREZID")

# Keep only rows from table without NAs
genes_24h <- genes_24h[is.na(genes_24h$Entrez.Gene)==FALSE,]

# Remove duplicated entries
genes_24h <- genes_24h[!duplicated(genes_24h$Entrez.Gene),]

# Filter genes with -0.5 log2 fold change
genes_24h_filter <- genes_24h %>%
  filter(crispr_1_LFC < -0.5)
  filter(crispr_2_LFC < -0.5)
  filter(run2_24h_LFC < -0.5)
  filter(merge_24h_1_LFC < -0.5)
  


```


```{r}
# Change Entrez IDs from numbers to characters
geneset <- as.character(genes_24h_filter$Entrez.Gene)

# This will show Entrez IDs only
head(geneset)

# This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

# dimensions - number of GO terms
dim(ego)

genes_24h_go <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)
# Copy and paste into Excel
```


```{r,fig.height=3, fig.width=8}
barplot (ego, showCategory = 10)
dotplot (ego, showCategory=10)
# can use simplify() to remove redundant terms, but can be slow eg barplot(simplify(ego))
```

We can look into the relation of the enriched GO terms using the `emapplot()` function. You will need to increase the height and width of the image so that the terms don't overlap.  

```{r, fig.height=12, fig.width=14}
emapplot(ego)
```

We can use cnetplot() to display the relation of specific genes within the GO terms.

```{r, fig.height=12, fig.width=14}
# logFC is optional, but if available will colour-code the nodes
logFC <- genes_24h_filter$crispr_1_LFC
# set name of object
names(logFC) <- genes_24h_filter$Entrez.Gene
cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
# show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
# size = number of genes, fold change = red for upregulated, green for downregulated
```


## 24 hours upregulated

```{r}
library(readr)
genes_24h <- read_csv("mcf_shared_genes_24h_up.csv")
genes_24h[1] <- NULL

# Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
genes_24h$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(genes_24h$SYMBOL), keytype="SYMBOL", column="ENTREZID")

# Keep only rows from table without NAs
genes_24h <- genes_24h[is.na(genes_24h$Entrez.Gene)==FALSE,]

# Remove duplicated entries
genes_24h <- genes_24h[!duplicated(genes_24h$Entrez.Gene),]

# Filter genes with 0.5 log2 fold change
genes_24h_filter <- genes_24h %>%
  filter(crispr_1_LFC > 0.5)
  filter(crispr_2_LFC > 0.5)
  filter(run2_24h_LFC > 0.5)
  filter(merge_24h_1_LFC > 0.5)
  
```


```{r}
# Change Entrez IDs from numbers to characters
geneset <- as.character(genes_24h_filter$Entrez.Gene)

# This will show Entrez IDs only
head(geneset)

# This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

# dimensions - number of GO terms
dim(ego)

genes_24h_go <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)
# Copy and paste into Excel
```


```{r,fig.height=3, fig.width=8}
barplot (ego, showCategory = 10)
dotplot (ego, showCategory=10)
# can use simplify() to remove redundant terms, but can be slow eg barplot(simplify(ego))
```

We can look into the relation of the enriched GO terms using the `emapplot()` function. You will need to increase the height and width of the image so that the terms don't overlap.  

```{r, fig.height=12, fig.width=14}
emapplot(ego)
```

We can use cnetplot() to display the relation of specific genes within the GO terms.

```{r, fig.height=12, fig.width=14}
# logFC is optional, but if available will colour-code the nodes
logFC <- genes_24h_filter$crispr_1_LFC
# set name of object
names(logFC) <- genes_24h_filter$Entrez.Gene
cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
# show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
# size = number of genes, fold change = red for upregulated, green for downregulated
```


## 7 days

```{r}
library(readr)
genes_24h <- read_csv("mcf_shared_genes_7d_down.csv")
genes_24h[1] <- NULL

# Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
genes_24h$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(genes_24h$SYMBOL), keytype="SYMBOL", column="ENTREZID")

# Keep only rows from table without NAs
genes_24h <- genes_24h[is.na(genes_24h$Entrez.Gene)==FALSE,]

# Remove duplicated entries
genes_24h <- genes_24h[!duplicated(genes_24h$Entrez.Gene),]

# Filter genes with -0.5 log2 fold change
genes_24h_filter <- genes_24h %>%
  filter(crispr_1_LFC < -0.5)
  filter(crispr_2_LFC < -0.5)
  filter(run2_24h_LFC < -0.5)
  filter(merge_24h_1_LFC < -0.5)
  


```


```{r}
# Change Entrez IDs from numbers to characters
geneset <- as.character(genes_24h_filter$Entrez.Gene)

# This will show Entrez IDs only
head(geneset)

# This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

# dimensions - number of GO terms
dim(ego)

genes_24h_go <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)
# Copy and paste into Excel
```


```{r,fig.height=3, fig.width=8}
barplot (ego, showCategory = 10)
dotplot (ego, showCategory=10)
# can use simplify() to remove redundant terms, but can be slow eg barplot(simplify(ego))
```

We can look into the relation of the enriched GO terms using the `emapplot()` function. You will need to increase the height and width of the image so that the terms don't overlap.  

```{r, fig.height=12, fig.width=14}
emapplot(ego)
```

We can use cnetplot() to display the relation of specific genes within the GO terms.

```{r, fig.height=12, fig.width=14}
# logFC is optional, but if available will colour-code the nodes
logFC <- genes_24h_filter$crispr_1_LFC
# set name of object
names(logFC) <- genes_24h_filter$Entrez.Gene
cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
# show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
# size = number of genes, fold change = red for upregulated, green for downregulated
```
