
---
title: "GO"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

# Gene ontology and gene set enrichment analysis
Pathway analysis using code from Rbio Session 4. Gene ontology using cluster Profiler https://yulab-smu.github.io/clusterProfiler-book/


```{r, include=FALSE}

library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)


#Go to Rbio.Rmd file to load full list, including installing packages etc. If having trouble, install individual packages using CRAN within R Studio. If still having trouble, then have to search Bioconductor for it.

```
## MCF-7 DMSO v SGC 7 days

### Filter data (process data)
The data contains double entries or NA entries. We need to remove these for downstream analysis.

```{r}

library(readr)
mcf_sig <- read_csv("results/MCF_DMSOvSGC_7D_sig.csv")

# Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are

mcf_sig$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(mcf_sig$Row.names), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
mcf_sig <- mcf_sig[is.na(mcf_sig$Entrez.Gene)==FALSE,]

#Remove duplicated entries
mcf_sig <- mcf_sig[!duplicated(mcf_sig$Entrez.Gene),]


```

### Gene Ontology (GO) over-representation test
We are going to perform an over-representation test for our genes in GO terms from org.Hs.eg.db

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(mcf_sig$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

#dimensions - number of GO terms
dim(ego)

mcf_GOgenes <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)
#Copy and paste into Excel

```
### Showing the ego object in graphical form

```{r,fig.height=4, fig.width=6}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

#can use simplify() to remove redundant terms, but can be slow eg barplot(simplify(ego))

```


We can look into the relation of the enriched GO terms using the `emapplot()` function. You will need to increase the height and width of the image so that the terms don't overlap.  

```{r, fig.height=8, fig.width=10}
emapplot(ego)

```

We can use cnetplot() to display the relation of specific genes within the GO terms.

```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC_mcf <- mcf_sig$log2FoldChange
#set name of object
names(logFC_mcf) <- mcf_sig$Entrez.Gene

cnetplot(ego, foldChange = logFC_mcf, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated


```

